CS 51 Final Project Technical Specification: Matrix Module and Simplex Algorithm---{\annot{Annotated!}}

### 2013-05-05 DraftSpec_Annotated.tex
Finished annotating Draft ie functional specs

% Uncomment this section if you wish to have a header.
\usepackage{fancyhdr} 
\pagestyle{fancy} 
\renewcommand{\headrulewidth}{0.5pt} % customise the layout... 
\lhead{CS51 Draft Spec} \chead{} \rhead{April 7, 2013} 
\lfoot{} \cfoot{\thepage} \rfoot{}

@@

\begin{document}
\title{CS 51 Final Project Draft Specification---\annot{Annotated!}}
\author{
Luis Perez $|$ luisperez@college.harvard.edu \\ 
Andy Shi $|$ andyshi@college.harvard.edu \\ 
Zihao Wang $|$ zihaowang01@college.harvard.edu \\ 
Ding Zhou $|$ dzhou@college.harvard.edu
}
\date{May 5, 2013}
\maketitle

%Put your stuff here
{\annot{Note for the reader: This is the annotated version of our CS 51 
final project specification. All annotations will appear in this color. Our changes to the actual modules and the finer points of implementation were too numerous to annotate (so there are no annotations in section 3)}}.

@@

Our project seeks to implement an efficient matrix library for Ocaml. The end
goal is to utilize this matrix library in conjunction with the simplex algorithm
to find solution to problems in linear programming. In order to accomplish this,
there will be multiple small checkpoints. We need to implement and efficient
matrix representation (so as to not use the naive implementation of lists of
lists). We also hope to provide a useful interface that allows the user to input
arbitrarily large matrices effectively (possible by mapping over the entries of
the matrix). One big challenge for the matrix library will be the implementation
of efficient algorithms for multiplication and row reduction. We expect the
multiplication to initially consists of the standard algorithm, but will
eventually grow into either block multiplication or Strassen {\annot{We were not
able to implement more sophisticated multiplication algorithms}}. We might even
implement some way to chose between the algorithms depending on the inputted
matrices, but that would not an essential feature {\annot{We were not able to do
this either}}.

The row reduction algorithm will certainly be one of the most fundamental
algorithms in our matrix library implementation. From our current understanding
of simplex, it seems like it will heavily rely on row reduction {\annot{The
simplex algorithm ended up relying on a different type of row reduction, so we
couldn't just plug in a matrix into our row reduction algorithm. We had to do a
lot of extra work to get simplex}}. The only algorithm we can come up with so
far is Gauss-Jordan, though we have found some algorithms which are more suited
for computation by computers with imprecise storage of floats {\annot{In fact,
we just had to be careful which element to choose as a pivot}}.

The simplex part of the problem is the main goal, but if we can get down the
matrix module, then it appears as if the simplex algorithm should be
straightforward {\annot{This was definitely not the case}}. Eventually, we wish
to implement a simplex module that allows the input of arbitrary linear
equations with linear constraints, and then we parse this information into the
corresponding matrix and finally perform the simplex algorithm on that matrix in
order to return a solution to the specified problem.

Other tentative goals are to implement matrices that can accept either
arbitrarily large numbers (bignums) or arbitrarily precise floats 
(\verb@int list * int stream@). It would also be interesting, though not
necessary, to implement multiple algorithms for each operation and choose from
them the one that is the best for a specific input data {\annot{We were not able
to implement any of this functionality}}.

@@

@@

\item Matrix norm 
{\annot{Did not implement}}

\item Matrix transpose
{\annot{Implemented as specified}}

\item Trace---This is a typical trace (sum of the entries on the diagonal)
{\annot{Implemented as specified}}

@@

\item Eigenvectors / Eigenvalues, calculated also by a method in Hubbard's book
(this might not be necessary to carry out the simplex algorithm and we might
abandon it if we can't figure out how to implement it in a timely manner) 
{\annot{Did not implement}}

@@

Our matrix module will actually be a functor that takes in an
\verb@ORDERED_AND_OPERATIONAL@ module (from ps4 and moogle) so we can have
matrices of ints, floats, or floats as streams.

@@

\begin{verbatim}
module type ORDERED_AND_OPERATIONAL
sig
  type t

@@

\begin{verbatim}
module type MATRIX = 
sig
  exception NonSquare

@@

\begin{verbatim}
module type simplex
sig
  type elt
  type linear_equation
  type constraint

@@

A helper for the row reduction which will find the minimum element and its
location.

>>>>

% Uncomment this section if you wish to have a header.
\usepackage{fancyhdr} 
\pagestyle{fancy} 
\renewcommand{\headrulewidth}{0.5pt} % customise the layout... 
\lhead{CS51 Draft Spec} \chead{} \rhead{May 5th, 2013} 
\lfoot{} \cfoot{\thepage} \rfoot{}

@@

\begin{document}
\title{CS 51 Final Project Functional Specification{\annot{:Matrix Module and Simplex Algorithm}}
---{\annot{Annotated!}}}
\author{
Luis Perez $|$ luisperez@college.harvard.edu \\ 
Andy Shi $|$ andyshi@college.harvard.edu \\ 
Zihao Wang $|$ zihaowang01@college.harvard.edu \\ 
Ding Zhou $|$ dzhou@college.harvard.edu
}
\date{May 5, 2013}
\maketitle

%Put your stuff here
{\annot{Note for the reader: This is the annotated version of our CS 51 
Final Project Functional Specifications. All annotations will appear in this color.}}.

@@

Our project seeks to implement an efficient matrix library for Ocaml. The end
goal is to utilize this matrix library in conjunction with the simplex algorithm
to find solution {\annot{Linear Programs}}. In order to accomplish this,
there will be multiple small checkpoints. We need to implement and efficient
matrix representation (so as to not use the naive implementation of lists of
lists){\annot{(Successfully accomplished. Our implementation uses array for efficienty)}}.
 We also hope to provide a useful interface that allows the user to input
arbitrarily large matrices effectively (possible by mapping over the entries of
the matrix){\annot{The user can input matrices with a simple text file composed of comma
seperated values}}. One big challenge for the matrix library will be the implementation
of efficient algorithms for multiplication and row reduction. We expect the
multiplication to initially consists of the standard algorithm, but will
eventually grow into either block multiplication or Strassen {\annot{We were not
able to implement more sophisticated multiplication algorithms, but it seems that the current
algorithm works fairly well}}. We might even implement some way to chose between the 
algorithms depending on the inputted matrices, but that would not an essential feature 
{\annot{This feature can still easily be implemented, but there simply was no time. We
already knew this, though, by the time we wrote our Technical Specs}}.

The row reduction algorithm will certainly be one of the most fundamental
algorithms in our matrix library implementation. From our current understanding
of simplex, it seems like it will heavily rely on row reduction {\annot{The
simplex algorithm ended up relying on a different type of row reduction, so we
couldn't just plug in a matrix into our row reduction algorithm. We had to do a
lot of extra work to get simplex. Though, many of the helper functions for row
reduction ended up being extremely useful for Simplex, so this was non as pointless
as some might thing}}. The only algorithm we can come up with so
far is Gauss-Jordan, though we have found some algorithms which are more suited
for computation by computers with imprecise storage of floats {\annot{In fact,
we just had to be careful which element to choose as a pivot. Futhermore, we
ended up implementing arbitrary precision, so this became a non-issue}}.

The simplex part of the problem is the main goal, but if we can get down the
matrix module, then it appears as if the simplex algorithm should be
straightforward {\annot{This was definitely not the case}}. Eventually, we wish
to implement a simplex module that allows the input of arbitrary linear
equations with linear constraints, and then we parse this information into the
corresponding matrix and finally perform the simplex algorithm on that matrix in
order to return a solution to the specified problem.{\annot{This is fully functionalt.
Input can be made in the form of a text file with a linear program. The algorithm
runs and not only returns the right values for simple cases, but also checks for
cases that might initially appear to be unsolvable but aren't, and for cases which
are unbounded.}}

Other tentative goals are to implement matrices that can accept either
arbitrarily large numbers (bignums) or arbitrarily precise floats 
(\verb@int list * int stream@). It would also be interesting, though not
necessary, to implement multiple algorithms for each operation and choose from
them the one that is the best for a specific input data {\annot{We were able to
implement arbitrary precisions and arbitrarily large sizes with the OCaml Nums
library}}.

@@

\item{\annot{Make a matrix from a text file. The user is now able to read in a
text file and create a matrix.}}

\item{\annot{For Information of further functionality, take a look at MatrixI.ml.}}

@@

\item Matrix norm 
{\annot{Did not implement since it was not necessary for Simplex.}}

\item Matrix transpose
{\annot{Implemented as specified.}}

\item Trace---This is a typical trace (sum of the entries on the diagonal)
{\annot{Implemented as specified.}}

@@

\item Eigenvectors / Eigenvalues, calculated also by a method in Hubbard's book
(this might not be necessary to carry out the simplex algorithm and we might
abandon it if we can't figure out how to implement it in a timely manner) 
{\annot{Did not implement since it was not necssary for Simplex.}}

@@

Our matrix module will actually be a functor that takes in an
\verb@ORDERED_AND_OPERATIONAL@ module (from ps4 and moogle) so we can have
matrices of ints, floats, or floats as streams. 
{\annot{This actually ended up being very helpful since at the end we we able 
to use OCaml's Num Library in order to achieve arbitrary precision and size.}}

@@

{\annot{For the updated view of this, please take a look at EltsI.ml}}
\begin{verbatim}
module type ORDERED_AND_OPERATIONAL
sig
  type t

@@

{\annot{For the updated view of this, please take a look at MatrixI.ml}}
\begin{verbatim}
module type MATRIX = 
sig
  exception NonSquare

@@

{\annot{For the updated view of this, please take a look at SimplexI.ml}}
\begin{verbatim}
module type simplex
sig
  type elt
  type linear_equation
  type constraint

@@

A helper for the row reduction which will find the minimum element and its
location.
{\annot{We ended up needing many more helper functions than those shown here.}}

### 2013-05-05 FinalSpec_Annotated.tex
Finished annotating Draft ie functional specs

% Uncomment this section if you wish to have a header.
\usepackage{fancyhdr} 
\pagestyle{fancy} 
\renewcommand{\headrulewidth}{0.5pt} % customise the layout... 
\lhead{CS51 Final Spec} \chead{} \rhead{April 14, 2013} 
\lfoot{} \cfoot{\thepage} \rfoot{}

@@

\begin{document}
\title{CS 51 Final Project Final Specification---{\annot{Annotated!}}}
\author{
Luis Perez $|$ luisperez@college.harvard.edu \\ 
Andy Shi $|$ andyshi@college.harvard.edu \\ 
Zihao Wang $|$ zihaowang01@college.harvard.edu \\ 
Ding Zhou $|$ dzhou@college.harvard.edu
}
\date{May 5, 2013}
\maketitle

%Put your stuff here
{\annot{To the reader: Annotations are again in this form}}.

@@

We have decided to reduce the amount of functions available to the outside
world. The matrix module will still contain the same functionality as before,
but only the functions necessary for the implementation of the Simplex algorithm
will be exposed (row-reduction, for example). The main focus of the project has
now shifted to having a working simplex algorithm. Other functions, such as
finding the eigenvalues of a matrix, while interesting, have been placed in our
“Cool Features” category. Furthermore, following the idea that Simplex is our
main focus, additional functionality such as parsing input, providing arbitrary
float precision and/or providing a matrix implementation with bignums has been
degraded in importance. All of our attention will be focused on creating a
working simplex algorithm. {\annot{All of this happened, except for the parsing input part. We were able to parse inputs from plaintext files}}.

>>>>

% Uncomment this section if you wish to have a header.
\usepackage{fancyhdr} 
\pagestyle{fancy} 
\renewcommand{\headrulewidth}{0.5pt} % customise the layout... 
\lhead{CS51 Final Spec} \chead{} \rhead{May 5th, 2013} 
\lfoot{} \cfoot{\thepage} \rfoot{}

@@

\begin{document}
\title{CS 51 Final Project Technical Specification: Matrix Module and Simplex Algorithm---{\annot{Annotated!}}}
\author{
Luis Perez $|$ luisperez@college.harvard.edu \\ 
Andy Shi $|$ andyshi@college.harvard.edu \\ 
Zihao Wang $|$ zihaowang01@college.harvard.edu \\ 
Ding Zhou $|$ dzhou@college.harvard.edu
}
\date{May 5, 2013}
\maketitle

%Put your stuff here
{\annot{Note for the reader: This is the annotated version of our CS 51 
technical project specification. All annotations will appear in this color. }}.

@@

We have decided to reduce the amount of functions available to the outside
world. The matrix module will still contain the same functionality as before,
but only the functions necessary for the implementation of the Simplex algorithm
will be exposed (row-reduction, for example{\annot{Row Reduction ended up not
being necessary. We have instead exposed a large set of library functions and,
additionally, some functions specific to Simplex that would not be exposed otherwise.
All of this is documented in MatrixI.ml}}). The main focus of the project has
now shifted to having a working simplex algorithm. Other functions, such as
finding the eigenvalues of a matrix, while interesting, have been placed in our
“Cool Features” category. Furthermore, following the idea that Simplex is our
main focus, additional functionality such as parsing input, providing arbitrary
float precision and/or providing a matrix implementation with bignums has been
degraded in importance. All of our attention will be focused on creating a
working simplex algorithm. {\annot{All of this happened, except for the parsing input part. We were able to parse inputs from plaintext files}}.
